# This is before doing my Monte Carlo simulation. This was built by filtering data, then logging everything but Eccentricity, then min-max into using a 0.97 threshold for Cosine Similarity



---
title: "MoFINAL"
author: "Mo"
date: "2026-06-04"
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: console
---

#Introduction
##Loading My Dataset
##In this section below, I'm importing data from the NASA Exoplanetary Archive. Each datapoint in the archive represents a confirmed exoplanet with measured features.
```{r load, message=FALSE, warning=FALSE}
# Packages list
# NOTE: "installr" removed - it's a Windows-only utility for updating R itself,
# it isn't used anywhere in this analysis, and it makes the install-check below
# noticeably slower to no benefit.
packages <- c("readxl", "psych", "igraph", "network",
              "ggnetwork", "visNetwork", "intergraph", "lsa",
              "data.table", "dplyr", "ggplot2", "ggfortify", "ggraph",
              "dbscan", "Rtsne", "uwot", "tidyr", "Matrix")

# Install missing packages and load them all
# installed.packages() rescans your whole library every single knit, which
# can itself take a while. Only run it once per session if you can help it.
new_packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new_packages)) install.packages(new_packages)
invisible(lapply(packages, library, character.only = TRUE))

#Read csv
NEA <- read.csv("cleanData.csv")

#Def target variables
target_vars <- c("pl_radj", "pl_bmassj", "pl_orbper", "pl_orbeccen", "pl_insol", "pl_dens")

# Make sure all our data is complete
NEA <- NEA[complete.cases(NEA[, target_vars]), ]
set.seed(123)
```

##In the next section, I fill in undefined entries and drop unwanted rows and columns within my excel file.
``` {r preprocess}
calc_flags <- c(
  "<a refstr=CALCULATED_VALUE href=/docs/composite_calc.html target=_blank>Calculated Value</a>",
  "<a refstr=CALCULATED_VALUE href=/docs/pscp_calc.html target=_blank>Calculated Value</a>"
)
NEA <- NEA[!(NEA$pl_radj_reflink %in% calc_flags), ]
NEA <- NEA[!(NEA$pl_bmassj_reflink %in% calc_flags), ]

# See Data
#View(NEA)
```

In the next section I define variables from the archive that represent common planetary features. Exploratory Analysis in R + Network Tutorial
What defines the similarity of a property between two planets (when do planets have a similar size or temperature)

#Louvain Section
## Cosine Similarity and Clustering
```{r 5D Cosine Similarity and Clustering}
# Set your threshold here
threshold_value <- 0.97

# 1. Prepare and Log-Transform the target variables
selected_columns <- NEA[, target_vars]

# Log-transform variables
log_vars <- c("pl_orbper", "pl_bmassj", "pl_radj", "pl_insol", "pl_dens")
log_vars <- intersect(log_vars, colnames(selected_columns))

selected_columns_logged <- selected_columns
for (v in log_vars) {
  selected_columns_logged[[v]] <- log10(selected_columns_logged[[v]])
}

# 2. Min-Max Scale the LOGGED columns (keeps values strictly between 0 and 1 for Cosine)
normalized_columns <- apply(selected_columns_logged, 2, function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
})

TRMOE_NEA <- t(normalized_columns)
colnames(TRMOE_NEA) <- NEA$pl_name
rownames(TRMOE_NEA) <- target_vars

# 3. Calculate 5D Cosine Similarity
cosine_TRMOE_ALL <- lsa::cosine(as.matrix(TRMOE_NEA))

# 4. Keep actual similarity weights, but drop edges below threshold
cosine_weighted <- cosine_TRMOE_ALL
cosine_weighted[cosine_weighted < threshold_value] <- 0
diag(cosine_weighted) <- 0

# 5. Build WEIGHTED network directly from a SPARSE matrix. This saves memory / makes it more efficient
cosine_sparse <- Matrix::Matrix(cosine_weighted, sparse = TRUE)

cosine_net <- graph_from_adjacency_matrix(
  cosine_sparse,
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)
V(cosine_net)$radius <- NEA$pl_radj * 4

# Clean up isolated nodes
isolated <- which(degree(cosine_net) == 0)
cosine_net_clean <- delete_vertices(cosine_net, isolated)
undirected_net <- simplify(cosine_net_clean)

# 6. Run Louvain (automatically respects edge weights!)
set.seed(123)
Louvain_Comm <- cluster_louvain(undirected_net)

# Lone Planets Data
isolated_planet_names <- V(cosine_net)$name[isolated]
lone_planets_table <- NEA[NEA$pl_name %in% isolated_planet_names, ]

# EXTRACT TO SEPARATE TABLES
cluster_assignments <- data.frame(
  pl_name = V(undirected_net)$name,
  Cluster = Louvain_Comm$membership
)

# Merge back
NEA_clustered <- merge(NEA, cluster_assignments, by = "pl_name")
cluster_tables <- split(NEA_clustered, NEA_clustered$Cluster)

# Calculate Modularity
mod_score <- modularity(Louvain_Comm)
print(paste("Network Modularity (Q):", round(mod_score, 4)))


#==========================================
#1. MODULARITY & HIERARCHICAL SUB-CLUSTERING
#==========================================
#Create Hierarchy Dataframe
hierarchy_df <- data.frame(
  pl_name = V(undirected_net)$name,
  Main_Cluster = Louvain_Comm$membership,
  Sub_Cluster = NA
)
#Loop for Sub-Clustering
set.seed(123)
for (i in unique(hierarchy_df$Main_Cluster)) {
  nodes_in_cluster <- which(hierarchy_df$Main_Cluster == i)
  subgraph <- induced_subgraph(undirected_net, nodes_in_cluster)
  sub_louvain <- cluster_louvain(subgraph)
  hierarchy_df$Sub_Cluster[nodes_in_cluster] <- paste0(i, ".", sub_louvain$membership)
}
#--- SAFELY ADD RADIUS: Uses matching to avoid breaking any downstream merges ---
hierarchy_df$Node_Radius <- NEA$pl_radj[match(hierarchy_df$pl_name, NEA$pl_name)]

# Attach the hierarchy attributes straight onto undirected_net so we never have
# to rebuild the graph or recompute cosine similarity again later.
v_idx <- match(V(undirected_net)$name, hierarchy_df$pl_name)
V(undirected_net)$Main_Cluster <- hierarchy_df$Main_Cluster[v_idx]
V(undirected_net)$Sub_Cluster  <- hierarchy_df$Sub_Cluster[v_idx]
V(undirected_net)$Node_Radius  <- hierarchy_df$Node_Radius[v_idx]

#==========================================
#2. SUB-CLUSTER CHEAT SHEET (PHYSICAL PROFILES)
#==========================================
#1. Merge the datasets together
NASA_SubClusters <- merge(NEA, hierarchy_df, by = "pl_name")
print(paste("Data check: Successfully merged", nrow(NASA_SubClusters), "planets!"))

#2. Run the physical profiling calculations using explicit dplyr tags
SubCluster_Profile <- NASA_SubClusters %>%
  dplyr::group_by(Sub_Cluster) %>%
  dplyr::summarize(
    Count = dplyr::n(),
    Med_Mass =  round(median(pl_bmassj, na.rm = TRUE), 3),
    Med_Radius = round(median(pl_radj, na.rm = TRUE), 3),
    Med_Period = round(median(pl_orbper, na.rm = TRUE), 2),
    Med_Eccen = round(median(pl_orbeccen, na.rm = TRUE), 3),
    Med_Insol = round(median(pl_insol, na.rm = TRUE), 1),
    Med_Dens =  round(median(pl_dens,   na.rm = TRUE), 3),
    .groups = "drop"
  ) %>%
  dplyr::arrange(Sub_Cluster)

#3. Clean printout
print(as.data.frame(SubCluster_Profile))


# MAIN CLUSTER IDENTIFICATION
main_cluster_summary <- NASA_SubClusters %>%
  group_by(Main_Cluster) %>%
  summarise(
    Count = n(),
    Med_Mass = median(pl_bmassj, na.rm = TRUE),
    Med_Radius = median(pl_radj, na.rm = TRUE),
    Med_Period = median(pl_orbper, na.rm = TRUE),
    Med_Eccen = median(pl_orbeccen, na.rm = TRUE),
    Med_Insol = median(pl_insol, na.rm = TRUE),
    Med_Dens = median(pl_dens, na.rm = TRUE),
    .groups = "drop"
  )

# View the summary table
print(main_cluster_summary)


#==========================================
#3. PREMIUM COLOR DICTIONARY
#==========================================
sub_cluster_colors <- c(
  #Cluster 1: Lava Worlds & Anomalies
  "1.1" = "#4A0004",
  "1.2" = "#991A1E",
  "1.3" = "#DF484E",

  #Cluster 2: Warm Saturns
  "2.1" = "#0B2545",
  "2.2" = "#1D528A",

  #Cluster 3: Rocky Core
  "3.1" = "#75460E",
  "3.2" = "#C98E2E",

  #Cluster 4: Mega-Hubs
  "4.1" = "#38125E",
  "4.2" = "#7F4AB0",

  #Cluster 5: Brown Dwarfs
  "5.1" = "#144018",
  "5.2" = "#3CA354",
  #Cluster 6: Deep Space Cold Worlds
  "6.1" = "#2B303A",
  "6.2" = "#8794A6"
)

# We reuse undirected_net from the clustering chunk above instead of rebuilding
# a second graph from scratch here. It already has Main_Cluster, Sub_Cluster,
# and Node_Radius attached as vertex attributes, and it's already had isolated
# nodes stripped out - this is used in both Viz's
global_strict_net <- undirected_net

stretched_layout <- layout_with_fr(
  global_strict_net,
  weights = E(global_strict_net)$weight,
  grid = "grid"
)
```
## Network visualization
# Subcluster Viz
```{r Subcluster visualization}
#==========================================
# 6. GLOBAL WEIGHTED NETWORK - SUB-CLUSTERS
#==========================================

# Reuse undirected_net and the exact layout from the main cluster plot
global_strict_net <- undirected_net

# Build the Visual using ggraph mapped to Sub_Cluster
final_sub_plot <- ggraph(global_strict_net, layout = stretched_layout) +
  geom_edge_link0(
    aes(
      width = weight,
      alpha = weight,
      color = as.character(node1.Sub_Cluster)
    )
  ) +
  scale_edge_width_continuous(
    range = c(0.1, 0.8),
    limits = c(threshold_value, 1.0)
  ) +
  scale_edge_alpha_continuous(
    range = c(0.03, 0.3),
    limits = c(threshold_value, 1.0)
  ) +
  scale_edge_color_manual(values = sub_cluster_colors, guide = "none") +
  geom_node_point(
    aes(fill = as.character(Sub_Cluster), size = Node_Radius),
    shape = 21,
    color = "black",
    stroke = 0.4,
    alpha = 0.90
  ) +
  scale_fill_manual(values = sub_cluster_colors, na.value = "gray80") +
  scale_size_continuous(range = c(0.8, 5.0), name = "Planet Radius (Rj)") +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),
    legend.title = element_text(face = "bold")
  ) +
  labs(
    title = "Exoplanet Network (Sub-Clusters)",
    subtitle = paste("Rad, Mass, Period, Eccen, Insol, Dens | (Threshold >=", threshold_value, ")"),
    fill = "Sub-Clusters",
    edge_width = "Similarity Strength",
    edge_alpha = "Similarity Strength"
  ) +
  guides(
    fill = guide_legend(override.aes = list(size = 5, shape = 21, color = "black"), ncol = 1),
    size = guide_legend(override.aes = list(fill = "gray40", shape = 21, color = "black"))
  )

# Save Plot
ggsave(paste0("6D_SubClusters_LogMinMax_", threshold_value, ".png"), plot = final_sub_plot, width = 13, height = 10, dpi = 300)

# Render plot in RStudio viewer
print(final_sub_plot)
```

```{r Viz 1.0 - Main Clusters, fig.width=12, fig.height=10}
#==========================================
# 5. GLOBAL WEIGHTED NETWORK - MAIN CLUSTERS (OPTIMIZED)
#==========================================

# 1. Define the main cluster color palette
main_cluster_colors <- c(
  "1" = "#991A1E", # Crimson Red (Rocky Worlds)
  "2" = "#1D528A", # Deep Blue (Neptunian Worlds)
  "3" = "#C98E2E", # Ochre Gold (Gas Giants)
  "4" = "#7F4AB0", # Amethyst Purple (Eccentric Orbits)
  "5" = "#3CA354", # Emerald Green (M-Dwarf Neptunes)
  "6" = "#8794A6"  # Slate Blue-Gray
)


# 2. Build the Visual using ggraph
final_plot <- ggraph(global_strict_net, layout = stretched_layout) +
  geom_edge_link0(
    aes(
      width = weight,
      alpha = weight,
      color = as.character(node1.Main_Cluster)
    )
  ) +
  scale_edge_width_continuous(
    range = c(0.1, 0.8),
    limits = c(threshold_value, 1.0)
  ) +
  scale_edge_alpha_continuous(
    range = c(0.03, 0.3),
    limits = c(threshold_value, 1.0)
  ) +
  scale_edge_color_manual(values = main_cluster_colors, guide = "none") +
  geom_node_point(
    aes(fill = as.character(Main_Cluster), size = Node_Radius),
    shape = 21,
    color = "black",
    stroke = 0.4,
    alpha = 0.90
  ) +
  scale_fill_manual(values = main_cluster_colors, na.value = "gray80") +
  scale_size_continuous(range = c(0.8, 5.0), name = "Planet Radius (Rj)") +
  theme_void() +
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5, size = 12, color = "gray40"),
    legend.title = element_text(face = "bold")
  ) +
  labs(
    title = "Exoplanet Network (Main Clusters)",
    subtitle = paste("Rad, Mass, Period, Eccen, Insol, Dens | (Threshold >=", threshold_value, ")"),
    fill = "Main Clusters",
    edge_width = "Similarity Strength",
    edge_alpha = "Similarity Strength"
  ) +
  guides(
    fill = guide_legend(override.aes = list(size = 5, shape = 21, color = "black"), ncol = 1),
    size = guide_legend(override.aes = list(fill = "gray40", shape = 21, color = "black"))
  )

# Save Plot
ggsave(paste0("6D_LogMinMax_", threshold_value, ".png"), plot = final_plot, width = 13, height = 10, dpi = 300)
```

``` {r Placeholder}
I looove science
```

## Boxplots for the Main Clusters
```{r MainClusters Boxplot}
# Step 1: Prepare the data, grouping purely by Main_Cluster
plot_data <- NASA_SubClusters %>%
  
  # We select only Main_Cluster and physical variables now
  select(Main_Cluster, pl_orbper, pl_bmassj, pl_radj, pl_insol, pl_dens, pl_orbeccen) %>%
  mutate(
    `Log10 Period\n(Days)` = log10(pl_orbper),
    `Log10 Mass\n(Jupiter masses)` = log10(pl_bmassj),
    `Log10 Radius\n(Jupiter radii)` = log10(pl_radj),
    `Log10 Insolation\n(Earth Flux)` = log10(pl_insol),
    `Log10 Density\n(g/cm^3)` = log10(pl_dens),
    `Eccentricity\n(Linear)` = pl_orbeccen  # Kept linear
  ) %>%
  select(-pl_orbper, -pl_bmassj, -pl_radj, -pl_insol, -pl_dens, -pl_orbeccen) %>%
  
  # Exclude only Main_Cluster from the pivot
  pivot_longer(cols = -Main_Cluster, names_to = "Feature", values_to = "Value") %>%
  mutate(Feature = factor(Feature, levels = c(
    "Log10 Period\n(Days)", "Log10 Mass\n(Jupiter masses)",
    "Log10 Radius\n(Jupiter radii)", "Log10 Insolation\n(Earth Flux)",
    "Log10 Density\n(g/cm^3)", "Eccentricity\n(Linear)"
  )))

# Define our cohesive palette mapping one color to each of the main clusters
main_cluster_colors <- c(
  "1" = "#991A1E", # (Rocky Worlds)
  "2" = "#1D528A", # (Neptunian Worlds)
  "3" = "#C98E2E", # (Gas Giants)
  "4" = "#7F4AB0", # (Eccentric Orbits)
  "5" = "#3CA354", # (M-Dwarf Neptunes)
  "6" = "#8794A6"  # If needed
)

# Step 2: Generate the faceted main cluster box plot
ggplot(plot_data, aes(x = as.character(Main_Cluster), y = Value, fill = as.character(Main_Cluster))) +
  geom_boxplot(
    alpha = 0.7,
    color = "black",
    outlier.shape = 1,       # Keeps outliers as hollow circles
    outlier.size = 1.5,
    outlier.colour = "black"
  ) +
  # Adds a solid black mean dot inside the boxes for mean
  stat_summary(fun = mean, geom = "point", shape = 20, size = 2.5, color = "black") +
  
  # Apply our custom main-cluster color palette
  scale_fill_manual(values = main_cluster_colors) +
  facet_wrap(~ Feature, scales = "free_y", nrow = 1) +
  theme_bw() +
  labs(
    title = "Physical Feature Distributions of Planetary Main Clusters",
    subtitle = "Faceted by Core Network Features",
    x = "Main Cluster ID",
    y = "Value (Log10 or Linear)"
  ) +
  theme(
    legend.position = "none", # No legend needed as x-axis labels represent the clusters directly
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10, face = "bold"), # Kept horizontal for clean reading
    strip.text = element_text(size = 9, face = "bold"),
    panel.grid.major.x = element_blank()
  )
```


## Network Metrics
```{r Network Metrics}
cat("\n==================================================\n")
cat("            1. GLOBAL NETWORK METRICS              \n")
cat("==================================================\n")

# Network Density: What percentage of all possible connections actually exist?
net_density <- igraph::edge_density(undirected_net)
cat("Global Network Density:", round(net_density, 5), "\n")

# Global Clustering Coefficient: Do similar exoplanets form tight, interconnected triads?
net_transitivity <- igraph::transitivity(undirected_net, type = "global")
cat("Global Clustering Coefficient (Transitivity):", round(net_transitivity, 4), "\n")


cat("\n==================================================\n")
cat("            2. NETWORK TOPOLOGY & SHAPE            \n")
cat("==================================================\n")

# Average Path Length: On average, how many similarity steps separate any two random planets?
avg_path_length <- igraph::mean_distance(undirected_net, directed = FALSE)
cat("Average Path Length (Degrees of Separation):", round(avg_path_length, 2), "steps\n")

# Network Diameter: What is the longest shortest-path across the entire exoplanet catalog?
net_diameter <- igraph::diameter(undirected_net, directed = FALSE)
cat("Network Diameter (Maximum Distance across Graph):", net_diameter, "steps\n")


cat("\n==================================================\n")
cat("            3. LOCAL NODE-LEVEL METRICS            \n")
cat("==================================================\n")

# A. Degree Centrality (Local Hubs: Planets with the highest volume of close neighbors)
local_degrees <- igraph::degree(undirected_net)
top_hubs <- head(sort(local_degrees, decreasing = TRUE), 5)

hubs_df <- data.frame(
  Planet_Hubs = names(top_hubs),
  Direct_Connections = as.vector(top_hubs)
)
cat("Top 5 Local Core Hubs (High Degree Centrality):\n")
print(as.data.frame(hubs_df))
cat("\n")

# B. Betweenness Centrality (Local Bridges: Planets that connect separate evolutionary regimes)
local_betweenness <- igraph::betweenness(undirected_net, directed = FALSE)
top_bridges <- head(sort(local_betweenness, decreasing = TRUE), 5)

bridges_df <- data.frame(
  Planet_Bridges = names(top_bridges),
  Bridge_Score = round(as.vector(top_bridges), 2),
  Connections = as.vector(igraph::degree(undirected_net)[names(top_bridges)])
)
cat("Top 5 Exoplanet Bridges (High Betweenness Centrality):\n")
print(as.data.frame(bridges_df))

# Filter the main dataset to pull the exact physical profiles of your 5 bridges
# Note: Fixed the trailing comma after pl_insol to prevent syntax crashes!
bridge_physical_profiles <- NEA_clustered %>%
  dplyr::filter(pl_name %in% names(top_bridges)) %>%
  dplyr::select(
    pl_name,
    pl_radj,       # Radius (Jupiter radii)
    pl_bmassj,     # Mass (Jupiter mass)
    pl_dens,       # Density
    pl_orbper,     # Orbital Period (Days)
    pl_orbeccen,   # Eccentricity
    pl_insol       # Insolation Flux
  )

cat("\n==================================================\n")
cat("      PHYSICAL TRAITS OF THE TOP 5 BRIDGES          \n")
cat("==================================================\n")
print(as.data.frame(bridge_physical_profiles))


# A. Degree Assortativity
net_assortativity <- igraph::assortativity_degree(undirected_net)
cat("Degree Assortativity Coefficient:", round(net_assortativity, 4), "\n")
if(net_assortativity > 0) {
  cat("  -> Positive: Hubs tend to connect to other Hubs (Rich-Club effect).\n\n")
} else {
  cat("  -> Negative: Hubs tend to connect to lower-degree fringe nodes.\n\n")
}

# B. Closeness Centrality (Top 5 "Baseline" Planets)
local_closeness <- igraph::closeness(undirected_net, normalized = TRUE)
top_closeness <- head(sort(local_closeness, decreasing = TRUE), 5)

closeness_df <- data.frame(
  Planet = names(top_closeness),
  Closeness_Score = round(as.vector(top_closeness), 4)
)
cat("Top 5 Highest Closeness Centrality (Most accessible from anywhere in the network):\n")
print(as.data.frame(closeness_df))
cat("\n")

# C. Eigenvector Centrality (Top 5 "Most Influential" Planets)
local_eigen <- igraph::eigen_centrality(undirected_net)$vector
top_eigen <- head(sort(local_eigen, decreasing = TRUE), 5)

eigen_df <- data.frame(
  Planet = names(top_eigen),
  Eigen_Score = round(as.vector(top_eigen), 4)
)
cat("Top 5 Highest Eigenvector Centrality (Connected to the most important Hubs):\n")
print(as.data.frame(eigen_df))

# ==========================================
# 12. ADVANCED TOPOLOGY: STRUCTURAL BOTTLENECKS
# ==========================================

cat("\n==================================================\n")
cat("      5. STRUCTURAL BOTTLENECKS & TIE STRENGTH     \n")
cat("==================================================\n")

# A. Articulation Points (Single Points of Failure)
# Identifies individual planets that, if removed, would snap the network into disconnected pieces.
art_points_idx <- igraph::articulation_points(undirected_net)
art_planets <- V(undirected_net)$name[art_points_idx]

cat("Articulation Points (Critical Evolutionary Bottlenecks):\n")
if(length(art_planets) > 0) {
  print(art_planets)
} else {
  cat("None found! Your network density is so high that there is no single point of failure.\n")
}

cat("\n")

# B. Edge Betweenness (Top 5 Most Critical Similarity Ties)
# Identifies the specific *connections* (not just planets) that bridge the largest structural gaps.
eb_scores <- igraph::edge_betweenness(undirected_net, directed = FALSE)
top_eb_idx <- order(eb_scores, decreasing = TRUE)[1:5]
top_edges <- igraph::E(undirected_net)[top_eb_idx]

# Extract the exact planet pairs holding these critical edges together
edge_list <- igraph::ends(undirected_net, top_edges)

eb_df <- data.frame(
  Planet_A = edge_list[, 1],
  Planet_B = edge_list[, 2],
  Edge_Betweenness_Score = round(eb_scores[top_eb_idx], 2)
)

cat("Top 5 Most Critical Similarity Ties (Edge Betweenness):\n")
print(as.data.frame(eb_df))
```
